Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-05-14 content.tex
Review manuscript, progress #46

Although many possible tests can be performed, 
we present here what we think are the five most important 
example research questions (ERQ) that \verb;pirouette; can answer, from simple questions to complex ones.
For each of these ERQs we will quantify the importance of a tree prior.

@@

@@

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model\_type} &
    \textbf{run\_if} &
    \textbf{do\_measure\_evidence} & 
    \textbf{inference model} \\ 
    \hline
    generative &
    always &
    FALSE &
    JC, strict, Yule \\
    \hline
  \end{tabular}
  \caption{
    Inference conditions and inference model for ERQ 1.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
  }
  \label{tab:RQ1}
\end{table}

@@

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_gen_experiment_explicit},
  caption = {
    Create the experiments to answer research question 1, 
    using explicit arguments.
  }
]
gen_experiment <- create_experiment(
  inference_conditions = create_inference_conditions(
    model_type = "generative", 
    run_if = "always"
  ), 
  inference_model = create_inference_model(
    tree_prior = create_yule_tree_prior(),
    clock_model = create_strict_clock_model(), 
    site_model = create_jc69_site_model()
  )
)
\end{lstlisting}

@@

\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single,
  label = {lst:create_gen_experiment},
  caption = {
    Create the experiments to answer research question 1.
  }
]
experiments <- list(gen_experiment)
\end{lstlisting}

The objects we just created must be bundled
in a \verb;pirouette; parameters structure,
together with error measurement parameters $\mathit{E}$ and
twinning parameters $\mathit{T}$. \giovanni{I suggest to refer to the twinning parameters only when strictly required and with the minimum possible amount of info provided, as we set it in such a way that the tree prior is imported from the inference model in an automatic way.}

We use the default error measurement
parameters, which uses a burn-in fraction of 10\% and the nLTT statistic to
measure the difference between phylogenies. For clarity,
we create this setup explicitly here:

@@

All these arguments are bundled
and checked by \verb;create_pir_params;:

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_pir_params},
  caption = Calling \texttt{create\_pir\_params}.
]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  error_measure_params = error_measure_params,
  twinning_params = NA
)
\end{lstlisting}

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_1/errors.png}
  \caption{
    ERQ 1: the inference error made 
    when the generative and inference model are the same.
    The vertical dashed line indicates the median error.
  }
  \label{fig:example_1}
\end{figure}

@@

In the previous example we selected the inference model $\mathit{I}$ to match 
with a known generative tree prior $\mathit{p_{G}^0}$.
However, a novel tree prior $\mathit{p_{G}}$, by our definition, is non-standard, 
and hence usually it is not part of a standard inference 
model (i.e. an inference model using a standard tree prior).
In such a case, the question is what standard tree prior should be 
used in the inference.
There are three ways to deal with this problem: hand-picking a tree
prior (as we did above), or picking the best of a set of tree priors, 
or to do both. 
Hand-picking a tree prior can be rational if one standard tree prior 
is more suitable than others for the inference, for example, when
the novel tree prior is most similar to a specific standard tree prior.
Instead of hand-picking, the best inference model from a list of candidate 
models can also be used in inference. Selecting the 'best' model in
\verb;pirouette; means selecting the model with the highest 
evidence (also known as marginal likelihood).

We show the procedure using \verb;pirouette; to answer the second research 
question: "What is the inference error made on a novel phylogeny when
using the best inference model, in comparison to a hand-picked model?

We will use the same tree as generated in \ref{lst:create_yule_tree}.

The first step is to simulate a DNA alignment from the 
given phylogeny. 
We will re-use the alignment parameters of the previous example 
as shown in Listing~\ref{lst:create_alignment_params}.

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model\_type} &
    \textbf{run\_if} & \textbf{do\_measure\_evidence} & 
    \textbf{inference model} \\ 
    \hline
    generative & always         & FALSE & JC, strict, Yule \\
    candidate  & best candidate & TRUE  & JC, strict, BD   \\
    candidate  & best candidate & TRUE  & JC, strict, CBS  \\
    ...        & ...            & ...   & ...              \\
    candidate  & best candidate & TRUE  & GTR, RLN, CCP    \\
    candidate  & best candidate & TRUE  & GTR, RLN, CEP    \\
    \hline
  \end{tabular}
  \caption{
    Inference conditions and inference model for ERQ 2.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
    BD: birth-death tree prior.
    GTR: GTR site model.
    RLN: relaxed log-normal clock model.
    CBS: coalescent Bayesian Skyline tree prior.
    CCP: coalescent constant-population tree prior.
    CEP: coalescent exponential-population tree prior.
  }
  \label{tab:RQ2}
\end{table}

In the second step, we deviate from the answer to ERQ 1 by providing a 
set of experiments $\{\mathit{X_{i}}\}_{i=1}^{40}$. All the experiments have 
the same inference conditions $\mathit{C_{i}} = \mathit{C}$ but different 
candidate inference models $\mathit{I_{i}}$. Only one of them will be used 
to run the actual full inference. According to the specified $\mathit{C}$, 
\verb;pirouette; will measure the evidence for each of them and select only 
the one with the highest evidence as the best candidate.

As model selection is commonly performed on the full list of available 
candidate models, \verb;pirouette; has a dedicated function for this choice:
\verb;create_all_experiments; will create a full set of 40 experiments,
containing the inference models of all combinations of 
4 site models, 2 clock models and 5 tree priors.

\begin{lstlisting}[
  language = R, 
  floatplacement = ht, 
  frame = single, 
  label = {lst:create_all_experiments},
  caption = {
    Create all 40 candidate experiments, except for the inference model of the generative model.
  }
]
cand_experiments <- create_all_experiments(
  exclude_model = gen_experiment$inference_model
)
\end{lstlisting}

\begin{lstlisting}[
  language = R, 
  floatplacement = ht, 
  frame = single, 
  label = {lst:create_all_experiments},
  caption = {
    Create a collection of experiments, with 1
    generative model, and 39 candidate models
  }
]
experiments <- list(
  list(gen_experiment),
  cand_experiments
)
\end{lstlisting}

For all the other steps we use the default setups. 
We can create the complete
\verb;pirouette; parameter set like this (which is the
same as Listing~\ref{lst:create_pir_params}, but using the defaults):

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_6/errors.png}
  \caption{
    ERQ 6: the inference error
    for both the generative and best candidate inference models.
  }
  \label{fig:example_6}
\end{figure}

@@

Up until now, we measured the baseline error of the \verb;pirouette; method,
by using the same standard tree prior in both the generation of 
the original tree, as well as in the inference.
This baseline error distribution serves as a control,
to quantify the impact of a new (non-standard) tree prior.
To measure that impact, we compare the inference error 
from the new tree prior to the baseline error,
by using a twin tree (see subsection~\ref{subsec:twinning}) for
both the generative and best candidate model.

The research question this example answers is:
"What is the inference error made from a phylogeny,
for both a generative model and a best candidate model, 
compared to the background noise?"

Most of the settings for this example are
the same as in ERQ 2: we will re-use the phylogeny as 
before (Listing~\ref{lst:create_yule_tree}), 
as well as the alignment parameters (Listing~\ref{lst:create_alignment_params}).
Also the experiments are the same as in 
Listing~\ref{lst:create_all_experiments}.

This time, however, we measure both the baseline and complete inference error.
We set the twinning parameters $\mathit{T}$ to create a twin tree,
as described in the 'Twinning' section \ref{subsec:twinning}.

Creating the twinning parameters is trivial with the default settings.
For clarity, however, we explicitly show the most
important arguments:

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_twinning_params},
  caption = Create the default twinning parameters.
]
twinning_params <- create_twinning_params(
  twin_model = "bd", 
  method = "random_tree"
)
\end{lstlisting}
\giovanni{We want to import the "twin\_model" from the inference model.}
\richel{Yes, that is an option, but according to Rampal not important enough
to add now}
\rampal{did I say that? I think it's OK not to do it in the paper, but perhaps it can be noted.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Interpretation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% RJCB will add: in six percent did the twin do worse, 
% see https://github.com/richelbilderbeek/pirouette_article/issues/41

In the five example research questions, 
we showed how to use \verb;pirouette; 
to quantify the importance of a tree prior in Bayesian phylogenetics.
In all ERQs we used the most simple generative tree prior possible,
which is the Yule tree prior.
In principle any other (more complex) generative tree prior 
can be tested, but we chose to provide the 
simplest (and fastest to run) examples.
All tests were performed on a single original tree, 
hence not providing us enough statistical power to 
evaluate the level of accuracy of the inference.

However, if the same procedure were repeated and performed on a distribution 
with a sufficient number of generative trees, 
it constitutes a quantitative assessment 
of the quality of the inference.
To give an idea of the insights \verb;pirouette; can give,
we pretend the results of the example research questions
to be obtained from thousands of replicates.
In that case, Figure~\ref{fig:example_1} would show 
the minimum error BEAST2 when the
generative model is known, which can serve as 
a minimum noise level assessment in a larger experiment.  
Figure~\ref{fig:example_2} shows that 
the best candidate model gives an error distribution similar to the unknown 
tree prior,
hinting that the candidate model would be just as good in inference.
Figure~\ref{fig:example_3} shows that 
the twin model gives an error distribution similar to the unknown tree prior,
hinting that the shape of the phylogeny with unknown tree prior
does not effect the error made by BEAST2.
Figure~\ref{fig:example_4} shows that 
the twin model and best candidate model give 
an error distribution similar to the unknown tree prior,
hinting that the shape of the phylogeny with unknown tree prior
does not effect the error made by BEAST2, 
and that the candidate model would be just as good in inference.

In conclusion, \verb;pirouette; can directly show 
the comparison between the results obtained 
adopting a new generative prior as compared to the error BEAST2 
would make when operating on a tree generated 
under a known, right and existing tree prior (the twin tree).
From such results an user can estimate whether or not a new tree prior, 
tailored on the generative process, is needed. 
If this were indeed the case, 
one could implement their own tree prior as a new module in BEAST2.

>>>>

Although many possible tests can be performed, 
we show the usage of \verb;pirouette; by a story that introduces
its features gradually, yet ending in quantifying the impact a tree prior has
in Bayesian inference.

@@

We show the usage of \verb;pirouette; by a story that introduces
its features gradually. First we measure the Bayesian inference error
when we have a phylogeny with a known and standard tree prior,
to get an idea of the baseline error.
Second, we'll additionally measure the inference error made by a
best candidate model, to establish that the true tree prior is indeed the best.
Lastly, we will seperate the baseline error and complete error in
a tree generated by an unknown or non-standard tree prior, 
to quantify the impact the tree prior has in Bayesian inference.

@@

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \texttt{model\_type} &
    \texttt{run\_if} &
    \texttt{do\_measure\_evidence} & 
    \texttt{inference model} \\ 
    \hline
    generative &
    always &
    FALSE &
    JC, strict, Yule \\
    \hline
  \end{tabular}
  \caption{
    Inference conditions and model.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
  }
  \label{tab:RQ1}
\end{table}

@@

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_gen_experiment_explicit},
  caption = {
    Create an experiment with the generative model,
    that will always be used in the actual inference, 
    using explicit arguments.
  }
]
generative_experiment <- create_experiment(
  inference_conditions = create_inference_conditions(
    model_type = "generative", 
    run_if = "always"
  ), 
  inference_model = create_inference_model(
    tree_prior = create_yule_tree_prior(),
    clock_model = create_strict_clock_model(), 
    site_model = create_jc69_site_model()
  )
)
\end{lstlisting}

@@

\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single,
  label = {lst:create_gen_experiment},
  caption = {
    Create the experiments. In this case, we use only
    an experiment with the generative model,
    that will always be used in the actual inference.
  }
]
experiments <- list(generative_experiment)
\end{lstlisting}

We also need to specify the error measurement parameters $\mathit{E}$.
Here we use the default error measurement
parameters, which uses a burn-in fraction of 10\% and the nLTT statistic to
measure the difference between phylogenies. For clarity,
we create this setup explicitly here:

@@

We now have create the alignment parameters,
our experiments and our error measure parameters.
These objects need to be bundled in a bigger
\verb;pirouette; parameter structure, 
using \verb;create_pir_params;:

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_pir_params},
  caption = Calling \texttt{create\_pir\_params}.
]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  error_measure_params = error_measure_params
)
\end{lstlisting}

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_1/errors.png}
  \caption{
    The inference error made 
    when the generative and inference model are the same.
    The vertical dashed line indicates the median error.
  }
  \label{fig:example_1}
\end{figure}

@@

In the previous example we selected the inference model $\mathit{I}$ to match 
with a known generative tree prior $\mathit{p_{G}^0}$.
However, a novel tree prior $\mathit{p_{G}}$, by our definition, 
is non-standard, and hence not part of a standard inference 
model (i.e. an inference model using a standard tree prior).
In such a case, the question is which standard tree prior should be 
used in the inference.
In this example, we re-use our hand-picked inference model,
but additionally pick a best inference model from a set of inference models.
With 'best' inference model, we mean the model with the highest 
evidence (also known as marginal likelihood), based on the alignment.
We show the procedure using \verb;pirouette; to answer a second research 
question: "What is the inference error made on a novel phylogeny when
using the best inference model, in comparison to a hand-picked model?

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model\_type} &
    \textbf{run\_if} & \textbf{do\_measure\_evidence} & 
    \textbf{inference model} \\ 
    \hline
    generative & always         & FALSE & JC, strict, Yule \\
    candidate  & best candidate & TRUE  & JC, strict, BD   \\
    candidate  & best candidate & TRUE  & JC, strict, CBS  \\
    ...        & ...            & ...   & ...              \\
    candidate  & best candidate & TRUE  & GTR, RLN, CCP    \\
    candidate  & best candidate & TRUE  & GTR, RLN, CEP    \\
    \hline
  \end{tabular}
  \caption{
    Inference conditions and model.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
    BD: birth-death tree prior.
    GTR: GTR site model.
    RLN: relaxed log-normal clock model.
    CBS: coalescent Bayesian Skyline tree prior.
    CCP: coalescent constant-population tree prior.
    CEP: coalescent exponential-population tree prior.
  }
  \label{tab:RQ2}
\end{table}

We will use the same tree as generated in \ref{lst:create_yule_tree},
as well as the same alignment parameters as shown in 
Listing~\ref{lst:create_alignment_params}.

We do need to specify a different set of experiments: we
need to state that we already have an experiment for
the generative model, as well as that we want all other inference
models to compete. We call the competing models 'candidate models',
as only the best will be used in the actual inference.
As model selection is commonly performed on the full list of available 
candidate models, \verb;pirouette; has a dedicated function for this choice:
\verb;create_all_experiments; will create a full set of 40 experiments,
containing the inference models of all combinations of 
4 site models, 2 clock models and 5 tree priors. All we need to add is
to exclude the inference model in the generative experiment:

\begin{lstlisting}[
  language = R, 
  floatplacement = ht, 
  frame = single, 
  label = {lst:create_all_experiments},
  caption = {
    Create all 40 candidate experiments, except for the inference model of the generative model.
  }
]
candidate_experiments <- create_all_experiments(
  exclude_model = generative_experiment$inference_model
)
\end{lstlisting}

We combine the generative and all candidate models into one set of experiments:

\begin{lstlisting}[
  language = R, 
  floatplacement = ht, 
  frame = single, 
  label = {lst:create_all_experiments},
  caption = {
    Create a collection of experiments, with 1
    generative model, and 39 candidate models
  }
]
experiments <- list(
  list(generative_experiment),
  candidate_experiments
)
\end{lstlisting}

We can now create the complete
\verb;pirouette; parameter set like this (which is the
same as Listing~\ref{lst:create_pir_params}, but using the defaults):

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_6/errors.png}
  \caption{
    The inference error
    for both the generative and best candidate inference models.
  }
  \label{fig:example_6}
\end{figure}

@@

Up until now, we measured the inference error on a tree
of which we knew the (standard) tree prior. The goal of \verb;pirouette;
is to measure the impact of a novel tree prior.
To do so, in this example, we'll use a tree generated by a non-standard
tree prior, of which we assume is closely related to the Yule tree prior.
To measure the impact of the tree prior, we measure both the
baseline error (that serves as a control) 
and the full inference error: their difference shows
the impact of the tree prior.
To do so, we use a twin tree (see Subsection~\ref{subsec:twinning}) for
both the generative and best candidate model.
The research question this example answers is:
"What is the inference error made from a phylogeny,
for both a generative model and a best candidate model, 
compared to the background noise?"

This time we start with a tree generated from an unknown 
diversification model, that has six taxa and a crown age of ten:
\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single, 
  label = {lst:unknown_phylogeny},
  caption = A phylogeny generated by an unknown diversification model.
]
phylogeny  <- ape::read.tree(
  text = "(((A:8, B:8):1, C:9):1, ((D:8, E:8):1, F:9):1);"
)
\end{lstlisting}
\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_6/true_tree.png}
  \caption{The tree derived from an unknown diversification process, 
    as created by listing~\ref{lst:unknown_phylogeny}.
  }
\end{figure}

Most of the other settings are the same as before:
we re-use the alignment parameters (Listing~\ref{lst:create_alignment_params}),
as well as the experiments (Listing~\ref{lst:create_all_experiments}).
This time, however, we enable twinning (see Subsection~\ref{subsec:twinning}),
by creating twinning parameters.
Creating these parameters is trivial with the default settings.
For clarity, however, we explicitly show the most
important arguments:

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_twinning_params},
  caption = Create the default twinning parameters.
]
twinning_params <- create_twinning_params(
  twin_model = "bd", 
  method = "random_tree"
)
\end{lstlisting}

@@

We showed how to use \verb;pirouette; 
to quantify the importance of a tree prior in Bayesian phylogenetics,
using the most simple generative tree prior possible.
In principle any other (more complex) generative tree prior 
can be tested, but we chose to provide the 
simplest (and fastest to run) examples.
All tests were performed on a single original tree, 
hence not providing us enough statistical power to 
evaluate the level of accuracy of the inference.

However, if the same procedure were repeated and performed on a distribution 
with a sufficient number of generative trees, 
it constitutes a quantitative assessment 
of the quality of the inference.

A twin tree does not always result in a lower error distribution,
due to the stochasticity in generating the twin tree.
For example, of the second tree used in the 'Usage' section,
94\% of all twin trees resulted in an error distribution with a lower median.

% Add: the effect of the alignment on this
% https://github.com/richelbilderbeek/pirouette_article/issues/45

To give an idea of the insights \verb;pirouette; can give,
we pretend the final result was to be obtained from thousands of replicates.
We ...

In conclusion, \verb;pirouette; can directly show 
the comparison between the results obtained 
adopting a new generative prior as compared to the Bayesian inference error
when operating on a tree generated under a known and standard tree 
prior (the twin tree).
From such results an user can estimate whether or not a new tree prior, 
tailored on the generative process, is needed. 
If this were indeed the case, 
one could rightly implement their novel tree prior as an addition to 
their favorite Bayesian inference tool.

@@
